Blum axioms and nondeterministic computation of functions

### 2016-05-18 paper.tex
paper.tex: Fiddle with the text so it fits in 10 pages.

\section{Our approach: Blum axioms and Gödel numberings}
\label{sec:our-approach}

@@

Since nondeterministic computation returns several values,
to associate it with functions we need to either restrict what things define a function
or to somehow extract a single value from all these branches of computation.
It is important to note that this is not the only approach available in the literature.
Complexity classes like $\ComplexityFont{NPVM}$
(see, for example, the paper of Selman~\cite[p.~359]{Selman1994})
are defined using \emph{multivalued functions},
which are allowed to return several values for a single input.
Another approach is to use \emph{function problems} associated to problems in $\NP$,
where the machine is required to return any certificate for the given instance,
or to reject the input if it is not in the language;
this yields the class $\FNP$~\cite[p.~229]{Papadimitriou1994}.
We will not consider these approaches here;
in particular, we will define $\FNP$ differently below.

@@

One important theorem that can be proven using acceptable Gödel numberings alone
is the recursion theorem
\cite[p.~181]{Rogers1987}.
It states that, if $\phi$ is any acceptable Gödel numbering
and $f$ is any total recursive function,
then there is some program $w_0$ such that $\phi_{w_0} = \phi_{f(w_0)}$.
That is,
if $f$ is any systematic transformation on programs,
there is a program $w$ (a fixed point for $f$) whose meaning under $\phi$ is unchanged.
The recursion theorem can be used,
for example, to show the existence of quines
(programs whose output are their own source codes)
in any Turing-complete programming language:
choose $f$ to be the function that, given a program $w$,
returns another program $f(w)$ that prints the string $w$ when run.
(For Turing machines, for instance,
we can encode the bits of $w$ in the transition table of $w$.)
Then, by the recursion theorem,
there is some program $w_0$ that is equivalent to its transformed version $f(w_0)$;
thus, $w_0$ already writes the string $w_0$, its own source code.
Therefore, any programming language has quines \cite[p.~227]{Kozen2006}.

@@

\vspace{6pt}
\begin{proposition}
    Define the partial function $f: \Sigma^* \times \Sigma^* \to \{0, 1\}$ by
    \begin{equation*}
        f(w, x) = \begin{cases}
            1, & \text{if the machine $w$ does not halt on $x$.} \\
            \text{undefined}, &\text{if $w$ halts on $x$.}
        \end{cases}
    \end{equation*}
    This function can be computed by a nondeterministic Turing machine under
    Hopcroft and Ullman's definition.
\end{proposition}
\vspace{-3pt}

@@

    If $w$ halts on $x$,
    there will be two halting branches of computation,
    each writing a different value in the tape,
    so, by definition, the function is not defined on this input.
    If $w$ never halts,
    then only the first branch will halt
    (and with $1$ written on the tape),
    so the function is defined on this input and its value is $1$.
\end{proof}

Therefore,
using Hopcroft and Ullman's definition,
we can compute some noncomputable functions,
violating the requirement~\ref{surjectiveness} of Gödel numberings.

We can try to fix this definition by forcing all branches to halt;
but then, as all branches are required to return the same value,
the machine will be (almost) deterministic.
Therefore, if there is a machine $M$ that computes the characteristic function
of the satisfiability problem in nondeterministic polynomial time,
we could simulate $M$ on a given input,
choosing (say) always the first option when confronted with nondeterminism.
If this branch of computation returns $1$,
then every branch returns $1$ and the input is satisfiable;
if this branch returns $0$, every branch returns $0$ and the input is unsatisfiable.
We thus could solve $\SAT$ in \emph{deterministic} polynomial time.

So, with this restriction,
we lose the apparent exponential speed-up in computation time
we have when using nondeterminism.

@@

Call $\FNP$ the class of all functions computable in polynomial time,
under our definition\footnote{
    As noted in section~\ref{sec:our-approach},
    this definition is different
    from the one given by Papadimitriou~\cite[p~229]{Papadimitriou1994}.
}.
A language $L$ is $\NP$-complete if both $L \in \NP$
and every language in $\NP$ reduces to $L$;
that is, for every language $L' \in \NP$,
there is a polynomial-time computable function $f$
such that, for every $x$,
\begin{equation*}
    x \in L' \text{ if and only if } f(x) \in L.
\end{equation*}
(We are using Karp reductions here \cite[p.~42]{AroraBarak2009}.)
If we rephrase in terms of the characteristic functions $1_L$ and $1_{L'}$,
of $L$ and $L'$, respectively,
we have $1_{L'}(x) = 1_L(f(x))$ for all $x$.
We will generalize specifically this equation to define $\FNP$-completeness.

@@

We can construct a $\FNP$-complete function based on the halting problem:
define $f : \{0, 1\}^* \times \mathbb N \to \{0, 1\}^*$
such that $f(w, n)$ is the lexicographically greatest value written by any branch of $w$
after running for $n$ steps.
Such functions are $\FNP$-complete
because they simulate Turing machines directly.

However, this definition is very rigid
and allow for few functions to be $\FNP$-complete.
For example, if $g$ is a function that returns
(say) a largest clique in a graph,
any candidate function for $\FNP$-completeness
must be able to output representation of graphs.
This restrict the class of $\FNP$-complete functions
to functions like the one mentioned above.

>>>>

\section{Our approach: Gödel numberings and Blum axioms}
\label{sec:our-approach}

@@

We are restricting ourselves to using single-valued functions,
but there are alternative approaches.
Complexity classes like $\ComplexityFont{NPVM}$
(see, for example, the paper of Selman~\cite[p.~359]{Selman1994})
are defined using \emph{multivalued functions},
which are allowed to return several values for a single input.
Another approach is to use \emph{function problems} associated to problems in $\NP$,
where the machine is required to return any certificate for the given instance,
or to reject the input if it is not in the language~\cite[p.~229]{Papadimitriou1994}.
We will not consider these approaches here.

@@

One important theorem that can be proven using acceptable Gödel numberings alone
is the recursion theorem
\cite[p.~181]{Rogers1987}.
It states that, if $\phi$ is any acceptable Gödel numbering
and $f$ is any total recursive function,
then there is some program $w_0$ such that $\phi_{w_0} = \phi_{f(w_0)}$.
That is,
if $f$ is any systematic transformation on programs,
there is a program $w$ (a fixed point for $f$) whose meaning under $\phi$ is unchanged.
The recursion theorem can be used,
for example, to show the existence of quines
(programs whose output are their own source codes)
in any Turing-complete programming language:
choose $f$ to be the function that, given a program $w$,
returns another program $f(w)$ that prints the string $w$ when run.
(For Turing machines, for instance,
we can encode the bits of $w$ in the transition table of $f(w)$.)
Then, by the recursion theorem,
there is some program $w_0$ that is equivalent to its transformed version $f(w_0)$;
thus, $w_0$ already writes the string $w_0$, its own source code.
Therefore, any programming language has quines \cite[p.~227]{Kozen2006}.

@@

\vspace{6pt}
\begin{proposition}
    Define the partial function $f: \Sigma^* \times \Sigma^* \to \{0, 1\}$ by
    \begin{equation*}
        f(w, x) = \begin{cases}
            1, & \text{if the machine $w$ does not halt on $x$.} \\
            \text{undefined}, &\text{if $w$ halts on $x$.}
        \end{cases}
    \end{equation*}
    This function can be computed by a nondeterministic Turing machine
    under Hopcroft and Ullman's definition.
\end{proposition}

@@

    If $w$ halts on $x$,
    there will be two halting branches of computation,
    each writing a different value in the tape,
    so, by definition, the function is not defined on this input.
    If $w$ never halts,
    then only the first branch will halt
    (and with $1$ written on the tape),
    so the function is defined on this input and its value is $1$.
\end{proof}
Therefore, under Hopcroft and Ullman's definition,
we can compute some noncomputable functions,
violating the requirement~\ref{surjectiveness} of Gödel numberings.

We can try to fix this definition by forcing all branches to halt;
but then, as all branches are required to return the same value,
the machine will be (almost) deterministic.
Therefore, if there is a machine $M$ that computes the characteristic function
of the satisfiability problem in nondeterministic polynomial time,
we could simulate $M$ on a given input,
choosing (say) always the first option when confronted with nondeterminism.
If this branch of computation returns $1$,
then every branch returns $1$ and the input is satisfiable;
if this branch returns $0$, every branch returns $0$ and the input is unsatisfiable.
We thus could solve $\SAT$ in \emph{deterministic} polynomial time.
So, with this restriction,
we lose the apparent exponential speed-up in computation time
we have when using nondeterminism.

@@

Call $\FNP$ the class of all functions computable in polynomial time,
under our definition\footnote{
    Note that this definition
    is different from the one given by Papadimitriou~\cite[p~229]{Papadimitriou1994}.
}.
A language $L$ is $\NP$-complete if both $L \in \NP$
and every language in $\NP$ reduces to $L$;
that is, for every language $L' \in \NP$,
there is a polynomial-time computable function $f$
such that, for every $x$,
\begin{equation*}
    x \in L' \text{ if and only if } f(x) \in L.
\end{equation*}
(We are using Karp reductions here \cite[p.~42]{AroraBarak2009}.)
If we rephrase in terms of the characteristic functions $1_L$ and $1_{L'}$,
of $L$ and $L'$, respectively,
we have $1_{L'}(x) = 1_L(f(x))$ for all $x$.
We will generalize specifically this equation to define $\FNP$-completeness.

@@

We can construct a $\FNP$-complete function based on the halting problem:
define $f : \{0, 1\}^* \times \mathbb N \to \{0, 1\}^*$
such that $f(w, n)$ is the lexicographically greatest value written by any branch of $w$
after running for $n$ steps.
Such functions are $\FNP$-complete
because they simulate Turing machines directly.
However, this definition is very rigid
and allow for few functions to be $\FNP$-complete;
the requirement of directly returning the output of the function $g$ above
restricts the class of $\FNP$-complete functions
to functions that perform simulations.
